---
title: "Results for the Real Data Analysis"
author: "David Kepplinger"
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(gridExtra)
library(stringr)
library(pense)
library(magrittr)
source('utilities-plotting.R')
RESULTS_PATH <- file.path('results', 'real_da')
CV_RESULTS_PATH <- file.path(RESULTS_PATH, 'cv')

FIGURE_PATH <- file.path('figures')

custom_pdf <- function (filename, width, height) {
  pdf_fun <- if (exists('cairo_pdf', mode = 'function')) {
    cairo_pdf
  } else {
    pdf
  }
  if (!dir.exists(FIGURE_PATH)) {
    dir.create(FIGURE_PATH)
  }
  fname <- file.path(FIGURE_PATH, paste0('glass-', basename(filename)))
  fname <- str_replace(fname, fixed('-1.pdf'), '.pdf')
  pdf_fun(filename = fname, width = width, height = height)
}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      dev = c('png', 'custom_pdf'),
                      fig.ext = c('png', 'pdf'))

## Load the data
requireNamespace('chemometrics')
requireNamespace('cellWise')
glass_data_env <- new.env(parent = emptyenv())
data(glass, package = 'chemometrics', envir = glass_data_env)
data(data_glass, package = 'cellWise', envir = glass_data_env)

glass_y_orig <- glass_data_env$glass[, 'P2O5']
glass_y <- log(glass_y_orig)
glass_x <- as.matrix(glass_data_env$data_glass[, 15:500])

## Helper functions
`%||%` <- function (x, y) {
  if (is.null(x)) {
    y
  } else {
    x
  }
}

extract_test_pred <- function (estimates, .id = NULL) {
  map2_dfr(estimates, names(estimates), function (x, suffix) {
    if (is.list(x)) {
      .id <- if (!is.null(.id)) {
        paste(.id, suffix, sep = '_')
      } else {
        suffix
      }
      if (is.null(x$test_pred)) {
        extract_test_pred(x, .id)
      } else {
        tibble(method = .id, 
               pred = x$test_pred, 
               zeta = x$zeta %||% NA_real_,
               alpha = x$alpha %||% NA_real_, 
               bdp = x$bdp %||% NA_real_,
               nnz = sum(abs(x$beta) > .Machine$double.eps))
      }
    } else {
      NULL
    }
  })
}

extract_cv_curve <- function (estimates, .id = NULL) {
  map2_dfr(estimates, names(estimates), function (x, suffix) {
    if (is.list(x)) {
      .id <- if (!is.null(.id)) {
        paste(.id, suffix, sep = '_')
      } else {
        suffix
      }
      if (is.null(x$cv_curve)) {
        extract_cv_curve(x, .id)
      } else {
        tibble(method = .id, 
               zeta = x$zeta %||% NA_real_, 
               bdp = x$bdp %||% NA_real_) %>% 
          cbind(x$cv_curve)
      }
    } else {
      NULL
    }
  })
}
```

## Nested CV results

```{r load-cv-results}
cv_results <- list.files(CV_RESULTS_PATH, full.names = TRUE) %>% 
  map_dfr(function (fname) {
    job_res <- readRDS(fname)
    
    tibble(seed = job_res$job$seed,
           fold = job_res$job$fold,
           obs_index = job_res$job$test_fold) %>% 
      cbind(extract_test_pred(job_res$estimates))
  }) %>%
  mutate(method = str_replace(method, fixed('cv_'), ''),
         cv_type = str_extract(method, '(se|min)$'),
         method = str_remove(method, '_(se|min)$'),
         method = recode_method(method)) %>% 
  filter(seed <= 50, seed >= 1,
         fold <= 6, fold >= 1)
```


```{r check-cv-results, warning=TRUE}
cv_results %>%
  summarize(`Missing seeds` = paste(setdiff(seq_len(50), unique(seed)), collapse = ', '), 
            .groups = 'drop') %>% 
  filter(nzchar(`Missing seeds`)) %>% {
    if (nrow(.) > 0L) {
      warning("One or more seeds are missing:")
      .
    } else {
      invisible(NULL)
    }
  }

cv_results %>%
  group_by(seed) %>% 
  summarize(`Missing folds` = paste(setdiff(seq_len(6), unique(fold)), collapse = ', '), 
            .groups = 'drop') %>% 
  filter(nzchar(`Missing folds`)) %>% {
    if (nrow(.) > 0L) {
      warning("One or more seeds are missing folds:")
      .
    } else {
      invisible(NULL)
    }
  }
```

### Number of non-zero coefficients

```{r}
cv_results %>% 
  filter(!is.na(method)) %>% 
  group_by(method, cv_type, seed, fold) %>% 
  summarize(nnz = nnz[[1L]]) %>% 
  group_by(method, cv_type) %>% 
  summarize(`Mean` = mean(nnz),
            `SD` = sd(nnz),
            `Median` = median(nnz),
            Minimum = min(nnz),
            `5% quantile` = quantile(nnz, 0.05),
            `95% quantile` = quantile(nnz, 0.95),
            Maximum = max(nnz)) %>% 
  arrange(`Mean`)
```

### Prediction accuracy

```{r compute-prediction_accuracy}
pred_acc <- cv_results %>% 
  filter(cv_type == 'se' | method == 'I-LAMM', !is.na(method)) %>%
  group_by(seed, method) %>%
  summarize(pred_err = tau_size(glass_y[obs_index] - pred), 
            pred_err_orig = tau_size(glass_y_orig[obs_index] - exp(pred)),
            .groups = 'drop')
```

I-LAMM has a median prediction accuracy of `r median(filter(pred_acc, method == 'I-LAMM')$pred_err)` (`r median(filter(pred_acc, method == 'I-LAMM')$pred_err_orig)` on the original scale).

```{r prediction_performance-log, fig.width=5, fig.height=3, out.width='66%', fig.align='center'}
pred_perf_plot <- pred_acc %>% 
  filter(method != 'I-LAMM') %>%
  pivot_longer(starts_with('pred_err'), names_to = 'scale', values_to = 'pred_err') %>%
  mutate(scale = recode(scale, 
                        pred_err = 'log scale [log % w/w]', 
                        pred_err_orig = 'original scale [% w/w]')) %>% 
  ggplot(aes(x = method, y = pred_err, 
             fill = method, color = method, alpha = method, shape = method)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.15, seed = 123), 
             alpha = 0.4,
             size = 1) +
  scale_fill_colorblind(guide = 'none', values = STYLE_COLOR) +
  scale_color_colorblind(guide = 'none',
                         values = c('Ada. MM' = '#444444',
                                    'Ada. PENSE' = '#444444',
                                    'Ada. EN' = '#444444',
                                    STYLE_COLOR[c('PENSE', 'MM', 'EN', 'I-LAMM')])) +
  scale_alpha_manual(guide = 'none',
                     values = c('Ada. MM' = 1,
                                'Ada. PENSE' = 1,
                                'Ada. EN' = 1,
                                'PENSE' = 0.25,
                                'MM' = 0.25,
                                'I-LAMM' = 1,
                                'EN' = .25)) +
  scale_shape_manual(guide = 'none', values = STYLE_SHAPE) +
  scale_x_discrete(labels = function (lbl) { str_replace_all(lbl, fixed(' '), '\n')}) +
  facet_wrap(~ scale, scales = 'free') +
  labs(y = expression(paste(tau, '-scale of prediction errors')), x = NULL, fill = NULL) +
  ggplot_theme(base_size = 11)

pred_perf_plot
```

## Fit to full data set
```{r load-full-estimates}
full_estimates <- readRDS(file.path(RESULTS_PATH, 'full_estimates.rds'))
```

```{r selected-predictors}
extract_nz_freqs <- function (estimate, name) {
  beta <- if (is.numeric(estimate$beta)) {
    nnz <- which(abs(estimate$beta) > 1e-12)
    sparseVector(estimate$beta[nnz], i = nnz, length = length(estimate$beta))
  } else {
    estimate$beta
  }
  tbl <- tibble(freq_index = beta@i,
                freq = colnames(glass_x)[beta@i])
  colnames(tbl)[[2]] <- name
  tbl
}

nz_freqs <- extract_nz_freqs(full_estimates$adapense$cv_se, 'Ada. PENSE') %>% 
  full_join(extract_nz_freqs(full_estimates$adamm$cv_se, 'Ada. MM')) %>% 
  full_join(extract_nz_freqs(full_estimates$adaen$cv_se, 'Ada. EN')) %>% 
  full_join(extract_nz_freqs(full_estimates$pense$cv_se, 'PENSE')) %>% 
  full_join(extract_nz_freqs(full_estimates$mm$cv_se, 'MM')) %>% 
  full_join(extract_nz_freqs(full_estimates$en$cv_se, 'EN')) %>% 
  full_join(extract_nz_freqs(full_estimates$ilamm, 'I-LAMM')) %>%
  arrange(freq_index) %>% 
  select(-freq_index)
```

Number of frequencies with non-zero coefficients:

```{r}
nz_freqs %>% 
  summarize(across(.fns = function (x) {
    sum(!is.na(x))
  }))
```

The selected frequencies are the following:

```{r}
nz_freqs %>% 
  mutate(across(everything(), coalesce, '—'))
```

```{r adapense_fitted-log, fig.width=2.6, fig.height=3.1, out.width="33%", fig.align='center'}
y_fit <- with(full_estimates$adapense$cv_se, intercept + as.numeric(glass_x %*% beta))
resid_scale <- mlocscale(glass_y - y_fit, 0.28)[['scale']]

outlier_cutoff <- 3
outlier_band_int <- resid_scale * outlier_cutoff
outlier_polygon <- data.frame(x = c(-10, -10, 50 - outlier_band_int,
                                    -10,  50, 50),
                              y = c(outlier_band_int - 10, 50,  50,
                                    -outlier_band_int - 10, 
                                    -outlier_band_int - 10, 
                                    -outlier_band_int + 50),
                              gr = c('u', 'u', 'u', 'l', 'l', 'l'), color = 'outer_band',
                              stringsAsFactors = FALSE)

fitted_vs_obs_data <- tibble(y = glass_y,
                             yfit = y_fit,
                             rel_resid = abs(y - y_fit) / resid_scale,
                             color = if_else(rel_resid > outlier_cutoff, 'outlier', 'inlier'))

fitted_vs_obs_plot <- ggplot(fitted_vs_obs_data, 
                             aes(x = y, y = yfit, color = color, alpha = color)) +
  geom_polygon(data = outlier_polygon, 
               aes(x = x, y = y, group = gr, color = color, fill = color),
               alpha = .1, linetype = '34') +
  geom_point(size = 2) +
  geom_abline(slope = 1, intercept = 0, color = 'gray20', lty = '22', size = 2/3) +
  scale_y_continuous(name = expression(paste('Fitted log-concentration of ', P[2], O[5], 
                                             ' [log % w/w]')),
                     breaks = seq(-4, 1, by = 1)) +
  scale_x_continuous(name = expression(atop(
    paste('Observed log-concentration of ', P[2], O[5]), 
    paste('\n[log % w/w]'))),
    breaks = seq(-4, 1, by = 1),
    minor_breaks = log(c(0.02, 0.06, 0.2, 0.6, 2, 6)),
    sec.axis = dup_axis(
      breaks = log(c(0.02, 0.06, 0.2, 0.6, 2, 6)),
      labels = c(0.02, 0.06, 0.2, 0.6, 2, 6),
      name = expression(paste('Observed concentration of ', P[2], O[5], ' [% w/w]')))) +
  scale_alpha_manual(guide = 'none', values = c('inlier' = 0.5, 'outlier' = 1,
                                                'outer_band' = 1)) +
  scale_color_colorblind(guide = 'none', values = c('inlier' = 'gray', 'outlier' = 'blue',
                                                    'outer_band' = 'lightblue')) +
  scale_fill_colorblind(guide = 'none', values = c('inlier' = 'gray', 'outlier' = 'blue',
                                                   'outer_band' = 'lightblue')) +
  coord_fixed(xlim = c(-4, 1), ylim = c(-3, 2)) +
  ggplot_theme(base_size = 11) +
  theme(panel.grid.minor.x = element_line(color = colorblind_cols('lightblue'), 
                                          size = rel(0.5), linetype = "dotted"),
        axis.title.x.bottom = element_text(hjust = 0.6))

fitted_vs_obs_plot
```

## Combined in a single plot

```{r combined, out.width='100%', fig.width=7, fig.height=3}
grid.arrange(pred_perf_plot + labs(title = "(a) Prediction accuracy"),
             fitted_vs_obs_plot + labs(title = "(b) Fitted vs. observed"),
             layout_matrix = matrix(c(1, 3, 2, 2), nrow = 2, ncol = 2),
             widths = grid::unit(c(0.66, 0.34), "npc"),
             heights = grid::unit(c(0.95, 0.05), "npc"))
```

## Appendix -- not part of the manuscript

### Verifying the grids of penalization levels

The grids for the estimates on the full data are well adjusted:

```{r app-lambda-grids-full-data, out.width='100%', dev = 'png'}
cbind(full_estimates$adapense$cv_se$cv_curve, method = 'Ada. PENSE') %>% 
  bind_rows(cbind(full_estimates$adamm$cv_se$cv_curve, method = 'Ada. MM')) %>% 
  bind_rows(cbind(full_estimates$mm$cv_se$cv_curve, method = 'MM')) %>% 
  bind_rows(cbind(full_estimates$pense$cv_se$cv_curve, method = 'PENSE')) %>% 
  ggplot(aes(x = lambda, y = cvavg, ymin = cvavg - 3 * cvse, ymax = cvavg + 3 * cvse)) +
  geom_pointrange() +
  scale_x_log10() +
  facet_wrap(~ method, ncol = 2, scales = 'free_x')
```

Similarly for the estimates in the nested CV, the vast majority of minima are above the smallest penalization.

```{r app-load-lambda-grids-cv}
cv_curves <- list.files(CV_RESULTS_PATH, full.names = TRUE) %>% 
  map_dfr(function (fname) {
    job_res <- readRDS(fname)
    
    tibble(seed = job_res$job$seed, 
           fold = job_res$job$fold) %>%
      cbind(extract_cv_curve(job_res$estimates))
  }) %>%
  mutate(method = str_replace(method, fixed('cv_'), ''))
```

```{r app-plot-lambda-grids-cv, out.width='100%', dev = 'png'}
cv_curves %>% 
  mutate(method = recode_factor(method,
                                pense_se = 'PENSE',
                                adapense_preliminary_min = 'PENSE-Ridge',
                                adapense_se = 'Ada. PENSE',
                                adamm_se = 'Ada. MM',
                                mm_se = 'MM',
                                .default = NA_character_)) %>% 
  filter(!is.na(method)) %>% 
  group_by(seed, fold, method, alpha, zeta) %>% 
  mutate(lambda_rel = lambda / max(lambda),
         lambda_best = if_else(cvavg == min(cvavg), lambda_rel, NA_real_),
         color = if_else(abs(min(lambda_rel) - lambda_best) < .Machine$double.eps, 
                         colorblind_cols('orange'), colorblind_cols('blue'))) %>% 
  ungroup() %>% 
  ggplot(aes(x = lambda_rel, y = cvavg, ymin = cvavg - cvse, ymax = cvavg + cvse,
             color = color,
             group = interaction(seed, fold))) +
  geom_line(alpha = 0.4, color = 'gray50') +
  geom_pointrange(aes(x = lambda_best)) +
  scale_color_identity() +
  scale_x_log10() +
  facet_wrap(~ method, scales = 'free_x')
```
